\Large{Research Seminar Software Science} \\ \huge{Simulation of monadic effects in Haskell}

### 2018-04-18 paper.tex
introduced multiple subsections for Monad section || included fail, exception and list monad

A straightforward consequence of laziness is the evaluation order of a Haskell program. Being demand-driven, it becomes impossible for a function call to produce any side effects. Therefore, Haskell is considered to be a \textit{pure language}. For example, a function of type $Char \rightarrow Char$ will neither write nor read any mutable variables. Nevertheless, side effects are without any doubt very convenient. Haskell's restricted side effects resulted in a sluggish input/output. For this reason, monadic I/O was created; thus by leading to a functional language with side effects. 
\section{Monads}
Monads originate from a section of mathematics called category theory. Luckily for us, it is not required to have prior knowledge in category theory in order to comprehend the application of monads in Haskell. 
Nevertheless, there exists a certain complexity when it comes to grasping to concept of a monad. From history, monads were initially adopted in Haskell to produce input and output operations. However, the capabilities of monads were not bound to just I/O. Additionally, monads can support multiple operations such as state, non-determinism, continuations, exceptions etc. \\ \\ 
The default monad is composed by three essential components: 
\begin{itemize}
    \item a type constructor
    \item a function return
    \item a bind operator ($\gg$=)
\end{itemize}
The \textit{return} function and the bind operator ($\gg$=) have the following types:

\begin{center}
   return   :: a $\rightarrow$ m a \\
   ($\gg$=) :: m a $\rightarrow$ (a $\rightarrow$ m b) $\rightarrow$ m b 
   \end{center}
In order for the above function definitions to hold, the monad laws need to be satisfied. There exist three monadic laws:

\begin{itemize}
    \item right unit \hspace{20mm} m $\gg$= return \textcolor{red}{=} m 
    \item left unit \hspace{20mm} return x $\gg$= f \textcolor{red}{=} f x   
    \item associative \hspace{15mm} (m $\gg$= f) $\gg$= g \textcolor{red}{=} m $\gg$= (\textbackslash x $\rightarrow$ f x $\gg$= g) 
\end{itemize}
According to the laws of right and left unit, the return serves as a collector of a value where no computation occurs. As the type of \textit{return} suggests, it takes a value of type \textit{a}, puts it in a special context \textit{m} and returns as a result \textit{m a}.
The Maybe monad has been selected to provide a concrete insight on how a monad functions.

>>>>

A straightforward consequence of laziness is the evaluation order of a Haskell program. Being demand-driven, it becomes impossible for a function call to produce any side effects. Therefore, Haskell is considered to be a \textit{pure language}. For example, a function of type $Char \rightarrow Char$ will neither write nor read any mutable variables. Nevertheless, side effects are without any doubt very convenient. Haskell's restricted side effects resulted in a sluggish input/output. For this reason, monadic I/O was created; thus by leading to a functional language with side effects. 
\section{Monads}
Monads originate from a section of mathematics called category theory. Luckily for us, it is not required to have prior knowledge in category theory in order to comprehend the application of monads in Haskell. 
Nevertheless, there exists a certain complexity when it comes to grasping to concept of a monad. From history, monads were initially adopted in Haskell to produce input and output operations. However, the capabilities of monads were not bound to just I/O. Additionally, monads can support multiple operations such as state, non-determinism, continuations, exceptions etc.
\subsection{Construction}
For now think of a monad as a way of chaining operations together. 
In order to have a monad, three components are essential to its construction: 
\begin{itemize}
    \item a type constructor
    \item a function return
    \item a bind operator ($\gg$=)
\end{itemize}
The \textit{return} function and the bind operator ($\gg$=) have the following types:

\begin{center}
   return   :: a $\rightarrow$ M a \\
   ($\gg$=) :: M a $\rightarrow$ (a $\rightarrow$ M b) $\rightarrow$ M b 
   \end{center}
In order for the above function definitions to hold, the monad laws need to be satisfied. There exist three monadic laws as presented below:

\begin{itemize}
    \item right unit \hspace{20mm} m $\gg$= return \textcolor{red}{=} m 
    \item left unit \hspace{20mm} return x $\gg$= f \textcolor{red}{=} f x   
    \item associative \hspace{15mm} (m $\gg$= f) $\gg$= g \textcolor{red}{=} m $\gg$= (\textbackslash x $\rightarrow$ f x $\gg$= g) 
\end{itemize}
According to the laws of right and left unit, the return serves as a collector of a value where no computation occurs. As the type of \textit{return} suggests, it takes a value of type \textit{a}, puts it in a special context \textit{m} and returns as a result \textit{m a}. Furthermore, the associative law guarantees that the bind operator ($\gg$=) is concerned only with the order of computation and not the manner they are nested together. The bind operation can be compared to a semicolon(;) by taking the output from the former step and feeding it to the next step.
\subsection{Flavors}

It is crucial to understand that there exist more than one kind of monad. As a result, the implementation of the bind function ($\gg$=) mentioned on the previous section differs on various monads.

\begin{itemize}
    \item \textbf{Fail Monad} - Consider the case that each step will \textit{return} either a success or a fail. The bind operator can be constructed in such a way that the next step will be executed only if the former step succeeds. Hence, there is no need to check if the previous step failed since the \textit{fail monad} will take care of this by automatically aborting in case of a former failing step.
    \item \textbf{Exception Monad} - Either returns a values or raises an exception. Since it is up to the developer to implement such a monad, he/she may decide to ignore the first exception and throw an exception only if a second one is found. 
    \begin{center}
        \textbf{data} Exc \textit{e a} = Exception \textit{e} $\vert$ OK \textit{a}
    \end{center}
    \item \textbf{List Monad} - Models non-deterministic computation and return a sequence of values. There exists the opportunity to program each step to return multiple values. The bind operator ($\gg$=) may be used to iterate over these values and pass each value to the upcoming step. Otherwise various loops would be needed to deal with all these results. As it can be seen, the bind operator of the list monad would be very beneficial by removing the complexity of adding loops to retrieve information. 
    \begin{center}
        \textbf{type} List \textit{a} = [\textit{a}]
    \end{center}
\end{itemize}

\subsection{Maybe}
The Maybe monad has been selected to provide a concrete insight on how a monad functions.
